Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-26 submission/manuscript.Rmd
shigeki results 2 edits

The testis contains both germ cells (meiotic cells, e.g. spermatogonia) and somatic cells (non-meiotic, e.g. Leydig cells) that support hormone production and germline functions. Intriugingly, many _Kdm5c_-KO testis and ovary enriched-DEGs have germline-specific functions, suggesting _Kdm5c_-KO cells fail to distinguish between the soma and germline. To test if this holds true for all _Kdm5c_-KO testis-biased DEGs, we first assessed their known functions through gene onotology analysis. We found _Kdm5c_-KO testis-enriched DEGs high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).

To further validate if these testis DEGs are truly germline genes, we then evaluated their expression in somatic versus germ cells within the testis. We first compared their expression in the testis with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^) that prevent the maturation of germ cells[@handelSertoliCellDifferentiation1979]. Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B). The only testis-enriched DEG that did not show considerable downregulation with germline depletion was _FK506 binding protein 6 (Fkbp6)_, the aforementioned testis gene that regulates piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012]. We then assessed testis-enriched DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018]. We found that while some testis-enriched DEGs were classified as specific markers for different germ cell types (e.g. spermatogonia, spermatocytes, round spermatids, and elongating spermatids) none marked somatic cells (Figure 2C). Together, these data suggest the somatic _Kdm5c_-KO brain aberrantly expresses germline genes.

We then wanted to more deeply characterize germline gene misexpression with _Kdm5c_ loss, but lacked a comprehensive list of mouse germline-enriched genes. To facilitate downstream analyses, we generated a curated list of germline-enriched genes using currently available RNA-seq datasets in Kit^W/Wv^ mice. Wild-type and Kit^W/Wv^ datasets included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015], as well as adult male testes[@muellerIndependentSpecializationHuman2013]. We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any adult wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource for assessing germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).

>>>>

The testis contains both germ cells (meiotic cells, e.g. spermatogonia) and somatic cells (non-meiotic, e.g. Leydig cells) that support hormone production and germline functions. To determine if _Kdm5c_-KO testis-enriched DEGs are somatic or germline genes, we first assessed their known functions through gene onotology analysis. We found _Kdm5c_-KO testis-enriched DEGs high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).

To further validate if these testis DEGs are truly germline genes, we then evaluated their expression in somatic versus germ cells within the testis. We first compared their expression in the testis with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^) that prevent the maturation of germ cells[@handelSertoliCellDifferentiation1979]. Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B). We then assessed testis-enriched DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018]. We found that while some testis-enriched DEGs were classified as specific markers for different germ cell types (e.g. spermatogonia, spermatocytes, round spermatids, and elongating spermatids) none marked somatic cells (Figure 2C). Together, these data demonstrate that the _Kdm5c_-KO brain aberrantly expresses germline genes, reflecting an erosion between somatic versus germline identity.

We then wanted to more deeply characterize germline gene misexpression with _Kdm5c_ loss, but lacked a comprehensive list of mouse germline-enriched genes. To facilitate downstream analyses, we generated a list of germline-enriched genes using currently available RNA-seq datasets in Kit^W/Wv^ mice. Wild-type and Kit^W/Wv^ datasets included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015], as well as adult male testes[@muellerIndependentSpecializationHuman2013]. We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any adult wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource to globally assess germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).
